Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-02-18 pkg/disttree/inst/draft/draft.Rnw
revised parts of the application section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2029 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 77d2a192a80e3385189f9dd53e5109389a89a824

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogenous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    paramer like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects where it would be challenging
to specify a distributional regression model like GAMLSS.

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas. More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
82~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{gneiting2005calibrated}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{gamboostLSS2016} using non-homogenous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
82~available regressors.

@@

However, the restriction to additive models remains. Therefore, a different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the next subsection of this article. In particular, the covariates can be used to separate the data into more homogeneous subgroups. Doing so after fitting a global model to the whole data set (without considering any covariates) and then fitting a local model to each subgroup should improve the model remarkably. This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.

@@

As explained in the introduction, distributional forests provide a new approach in probabilistic forecasting by embedding the ideas of flexible parametric modeling into the framework of tree-structured models. In that way this novel methodology offers additional features that could be particularly valuable in probabilistic forecasting but have not yet been applied in this kind of settings. To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities. Their results are compared to those of other zero-censored Gaussian models including a standard ensemble model output approach and two GAMLSS models. For one of the GAMLSS models the regressors have been selected and interactions have been specified in advance based on subject-matter knowledge and experience from meteorology. The other one is based on a computationally more demanding boosting approach.

@@

%\newpage
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
\label{covariates}
\end{minipage} 
%}
\end{table}

@@

\item A GAMLSS where for each distribution parameter the selection of covariates and the specification of interactions is done in advance, both based on meteorological expert knowledge following \cite{Stauffer+Umlauf+Messner:2017} (prespecified GAM). Among the available 82 regressors, 9 where selected for the location parameter and 6 for the scale parameter. Additionally, for each distribution parameter one interaction term is specified in which convective availbale energy is interacting with total precipitation (with the ensemble mean for the location parameter and with the ensemble standard deviation for the scale parameter). 
%In this terms an ensemble mean of 6 hour sums of convective available energy is interacting with an ensemble mean or standarad deviation of 6 hour sums of total precipitation.  
These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation. By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information (and is therefore expected to profit from this advantage).
\item A boosted GAMLSS \citep{gamboostLSS2016} using non-homogenous boosting
\citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM). The idea of this (boosting) approach is to iteratively minimize a loss function where in each step only the coefficient leading to the highest improvement of the current model is updated. As the initial coefficients are all zero, only the most important covariates are included with nonzero coefficients while less relevant variables are ignored if the procedure is stopped before convergence.
\end{itemize}

@@

%%%<TH> always us the term "distributional forest"
As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates. A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable. In a forest model though, for each node in each of the trees a subset of these covariates is chosen randomly as possible split variables to select one from. Moreover, each tree is built on a resampled version of the learning data. For these reasons the choices of split variables vary over the trees of a forest making it difficult to analyze the impact of covariates just by investigating the different trees. To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied. Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 yearz are listed in Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation is the most influential variable. In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.

>>>>

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{breiman1984classification}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{Devel_treeforest}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{breiman2001random}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{Devel_treeforest}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{Devel_parmod} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogeneous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    parameter like location, scale, or shape (Figure~\ref{Devel_treeforest}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{Devel_treeforest},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects where it would be challenging
to specify a distributional regression model like GAMLSS.

In weather forecasting, these properties are appealing in mountainous regions
and complex terrain where small-scale effects are not well-resolved in the NWP
outputs. As these are computed at a coarser scale, there may be abrupt local
effects within a single NWP grid cell. To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas. More specifically, a zero-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of no precipitation on a
given day (see Figure~\ref{axams_onestation}). Forecasts for July are
established based on data from the same month in the years 1985--2012 along with
82~covariates derived from a wide range of NWP ensemble outputs. As
Figure~\ref{axams_onestation} shows, the station-wise models yield a full
distributional forecast for each day -- here for the same day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{ProbabilisticForecasting} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach (EMOS, \citealp{gneiting2005calibrated}), a GAMLSS
with regressors prespecified based on meteorological expertise following
\cite{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{gamboostLSS2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection from all
82~available regressors.

@@

However, the restriction to additive models remains. Therefore, a different approach on how to include covariates that also allows for non-additive effects and interactions is considered in the following. In particular, the covariates can be used to separate the data into more homogeneous subgroups. Doing so after fitting a global model to the whole data set (without considering any covariates) and then fitting a local model to each subgroup should improve the model remarkably. This procedure of splitting a data set into subgroups can be done by applying a tree algorithm.

@@

%% TO DO: valueable features, why relevant in weather forecasting, what was missing here before? (complex terrain, many covariates, all kinds of effects could appear, interactions unknown)

In weather forecasting, various factors can make it challenging to build a well fitting model. Complex physical processes in the atmosphere depend on possibly rapidly changing circumstances. Often it is hard to tell which of these circumstances have a relevant impact on the quantity of interest and what type of effect that can be or how they might interact. A profound basis of information can be obtained from numerical weather prediction (NWP) models as they provide predictions for a large number of meteorological quantities. However, while including too many variables in a model can lead to overfitting, excluding variables in advance can result in a loss of valuable information. Therefore, the task of selecting variables and specifying interactions forms the basis of building an adequate model. Moreover, this model has to be able to deal with a wide range of effects as abrupt changes can appear as well as smooth transitions. Especially if the considered region includes complex terrain, a large grid spacing of NWP models can lead to abrupt local effects within a single NWP grid cell.

As explained in the introduction, distributional forests provide a new approach in probabilistic forecasting by embedding the ideas of flexible parametric modeling into the framework of tree-structured models. The ability to deal with a wide variety of effects and an automatic variable selection and detection of interactions are features provided by distributional forests that could be particularly valuable in probabilistic (weather) forecasting in order to tackle the above mentioned challenges.
To show their strengths in practice, distributional forests are now applied to probabilistic precipitation forecasts in a mountainous region based on a large number of numerical weather prediction quantities. Their results are compared to those of other zero-censored Gaussian models including a standard ensemble model output approach and two GAMLSS models. For one of the GAMLSS models the regressors have been selected and interactions have been specified in advance based on subject-matter knowledge and experience from meteorology. The other one is based on a computationally more demanding boosting approach.

@@

%\newpage
\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation            & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $\frac{1}{1.6}$) &    & ensemble std. deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $\frac{1}{1.6}$)&    & \qquad all for 6--30 UTC\\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 UTC\\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimal values,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: preciptable water          &   & ensemble std. deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimal values,\\
\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 UTC \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 UTC\\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 UTC\\
\hline
\end{tabular}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.}
\label{covariates}
\end{minipage} 
%}
\end{table}
\fixme{dswrf\_mean/sprd\_min not included in formula but in the 82 covariates}

@@

\item A GAMLSS where for each distribution parameter the selection of covariates and the specification of interactions is done in advance, both based on meteorological expert knowledge following \cite{Stauffer+Umlauf+Messner:2017} (prespecified GAM). Among the available 82 regressors, 9 where selected for the location parameter and 6 for the scale parameter. Additionally, for each distribution parameter one interaction term is specified in which convective available energy is interacting with total precipitation (with the ensemble mean for the location parameter and with the ensemble standard deviation for the scale parameter). 
%In this terms an ensemble mean of 6 hour sums of convective available energy is interacting with an ensemble mean or standarad deviation of 6 hour sums of total precipitation.  
These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation. By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information (and is therefore expected to profit from this advantage).
\item A boosted GAMLSS \citep{gamboostLSS2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} for an automatic selection of covariates (boosted GAM). The idea of this (boosting) approach is to iteratively minimize a loss function where in each step only the coefficient leading to the highest improvement of the current model is updated. As the initial coefficients are all zero, only the most important covariates are included with nonzero coefficients while less relevant variables are ignored if the procedure is stopped before convergence.
\end{itemize}

@@

%%%<TH> always us the term "distributional forest"
As shown in Table~\ref{model_specification} the distributional forest includes all 82~covariates. A single tree model would allow to easily analyze which covariates are used as split variables in the nodes and hence have an impact on the response variable. In a forest model though, for each node in each of the trees a subset of these covariates is chosen randomly as possible split variables to select one from. Moreover, each tree is built on a resampled version of the learning data. For these reasons the choices of split variables vary over the trees of a forest making it difficult to analyze the impact of covariates just by investigating the different trees. To get an overview of which covariates have the most influential effect on the response variable a standard measurement of variable importance for forest models has been applied. Based on the mean decrease in accuracy regarding the CRPS the 10 leading covariates of the above described distributional forest for station Axams learned on the first 24 years are listed in Figure~\ref{varimp}. As expected, the numerical forecast for total precipitation is the most influential variable. In particular, five variations of this variable are among the top ten of the list of covariates ordered by variable importance.
